---
title: Detailed Analysis
author: Lisa Hu, Niek Scholten
output:
  pdf_document:
    latex_engine: xelatex
---
```{r setup, include = FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(error = FALSE)
knitr::opts_chunk$set(results = "hide")

packages <- c("deSolve", "ggplot2", "formatR", "scales", "ggpubr")
invisible(lapply(packages, library, character.only = T))
```

# Data loading
```{r data-loading}
## Read the data
MPL_data <- read.csv("MPL.csv", na.strings = "NA")
## Rename the receptor
names(MPL_data)[4:5] <- c("R.m0", "R.0")

## Create the medians subsets
medians <- aggregate(MPL_data[,c("MPL_conc","R.m0","R.0")],
                     list(MPL_data$dose, MPL_data$time), median, na.rm=T)
names(medians)[1:2] <- c("dose","time")
```

# Assignment 1
## Implement the model
Following experiments with methylprednisolone in rats, a set of values for the model is defined:

| Parameter | Value   | Unit     | Explanation                                                               |
|-----------|---------|----------|---------------------------------------------------------------------------|
| k~s_Rm~   | 2.90    | fmol/g   | Zero-order rate constant: GR mRNA synthesis                               |
| IC~50_Rm~ | 26.2    | fmol/mg  | Concentration DR(N) that inhibits mRNA synthesis (50%)                    |
| k~on~     | 0.00329 | L/nmol/h | Second-order rate constant: forming MPL-receptor complex                  |
| k~T~      | 0.63    | 1/h      | First-order rate constant: translocation of receptor (cytosol -> nucleus) |
| k~re      | 0.57    | 1/h      | First-order rate constant: recovery of receptor (nucleus -> cytosol)      |
| R~f~      | 0.49    | -        | Fraction free receptor that gets recycled                                 |
| k~d_R~    | 0.0572  | 1/h      | First-order rate constant: breakdown of receptor                          |
| k~d_Rm~   | 0.612   | -        | First-order rate constant: breakdown of GR mRNA                           |
| k~s_R~    | 3.22    | -        | First-order rate constant: production of receptor                         |
| D         | ~       | nmol/L   | Concentration MPL (Calculated with molecular weight = 374.471             |
Table: Parameter values for MPL

| Parameter            | Value | Unit    | Explanation                  |
|----------------------|-------|---------|------------------------------|
| R~m0~ (mRNA)         | 4.74  | fmol/g  | Concentration mRNA~R~        |
| R~0~ (Free_receptor) | 267   | fmol/mg | Concentration free receptors |
| DR                   | 0     | fmol/mg | MPL in the cytosol           |
| DR(N)                | 0     | fmol/mg | MPL in the nucleus           |
Table: Initial values for MPL

\newpage

To create the plots for the model, functions can be used:
```{r create-functions-a1}
## Model
grd_model <- function(t, y, parms){
  with(as.list(c(parms, y)),{
    delta.R.m0 <- k.s_Rm * (1 - ( DRN / (IC.50_Rm + DRN) ) ) - k.d_Rm * R.m0
    delta.R.0 <- k.s_R * R.m0 + R.f * k.re * DRN - k.on * D * R.0 - k.d_R * R.0
    delta.DR <- k.on * D * R.0 - k.T * DR
    delta.DRN <- k.T * DR - k.re * DRN
    return( list( c(delta.R.m0, delta.R.0, delta.DR, delta.DRN ) ) )
  })
}

## Data frame with the different units and titles of the data
data.values <- data.frame(
        name = c("R.m0", "R.0", "DR", "DRN"),
        ylabel = c("mRNA", "Free receptor", "Receptor complex", "Receptor complex"),
        unit = c("(fmol/g)", "(fmol/mg)", "(fmol/mg)", "(fmol/mg)"),
        title = c("mRNA", "Free receptor", "Receptor complex (cytosol)",
                   "Receptor complex (nucleus)") )
rownames(data.values) <- data.values$name

## Function that calculates D
calc.D <- function(dose){
  return( (dose * 1000)/374.471 )
}

## Function to create a plot depending on wanted data
create.plot1 <- function(plot.value, simulation, median.data, model.data){
  # y.val inserts the plot.value for the corresponding row of data.values
  y.val <- data.values[plot.value,]
  # The plot
  plt <- ggplot(data = simulation, mapping = aes(x = time, y = !!sym(y.val$name) ) ) +
            # Points and lines
            geom_point(size = 1, shape = 1, aes(color = "Experiment") ) +
            geom_line(data = median.data, aes(color = "Median")) +
            geom_line(data = model.data, aes(color = "Model")) +
            # Labels
            labs(x = "Time", y = paste(y.val$ylabel, y.val$unit), title = y.val$title) +
            theme(legend.position = "bottom") +
            # Legend data
            scale_colour_manual(values = c("black", "red", "black"),
                                limits = c("Experiment", "Median", "Model") ) +
            # Legend correction
            guides(color = guide_legend(title = "",
                                        override.aes = list(linetype = c(1, 1, NA),
                                                            shape = c(NA, NA, 1)) ) )
  return(plt)
}

## Function arranges plots and legend
arrange.plots <- function(plots, title){
  my.grid <- ggarrange(plotlist = plots, ncol = 2, nrow = length(plots)/2,
                       common.legend = TRUE, legend = "bottom")
  print( annotate_figure(my.grid, top = text_grob(title) ) )
}

## Function joins other functions with data
main <- function(dose){
  # Change D depending on dose
  params$D <- calc.D(dose)
  # The different datas
  simulation <- subset(MPL_data, MPL_data$dose == 0 | MPL_data$dose == dose)
  median.data <- subset(medians, medians$dose == 0 | medians$dose == dose)
  # Run the model
  model.data <- as.data.frame( ode(times = times, y = state, parms = params,
                                   func = grd_model, method = "euler"))
  # Create the plots
  plots <- lapply(c("R.m0", "R.0"), create.plot1, simulation, median.data, model.data)
  # Arrange the plots
  arrange.plots(plots, sprintf("Graphs for %s mg drug", dose))
}
```

\newpage

The experiment followed rats for 7 days under constant infusion of the drug: 0.1 or 0.3 mg~drug~/kg~rat~/h.
With this information, models can be created:
```{r model, fig.height = 3}
## Set the data for the model
params <- c(k.s_Rm = 2.90, IC.50_Rm = 26.2, k.on = 0.00329, k.T = 0.63, k.re = 0.57,
            R.f = 0.49, k.d_R = 0.0572, k.d_Rm = 0.612, k.s_R = 3.22, D = 0)

state <- c(R.m0 = 4.74, R.0 = 267, DR = 0, DRN = 0)

times <- seq(0, 168, by = 1)

## Check the doses
doses <- unique(MPL_data$dose)
doses <- doses[!doses == 0]
sapply(doses, main)
```

\newpage

## Questions

* [1] Why is it best practice to plot the median for the experimental data?

> The median does not change with huge outliers, so the data is more reliable. If there is a significant difference, it will show.

* [2] How do the results of the simulations depend on the dose and concentration of the drug?

> The dose can influence the steady state of the different concentrations. The shape of the median lines is generally unaffected.

* [3] Are the results of the model in line with experimental data?

> Yes, to some extent. The values that the experiment and the model end on is similar, but the values between the start and the end fluctuate a bit. The graphs of the free receptor concentration show that the experimental data conform with the model. In the mRNA graphs, there are some deviations from the model around Time = 25.

\newpage

# Assignment 2
## No Auto-Regulation Glucocorticoid Receptor

```{r nodrugs, fig.height = 3}
## Create function for no drugs
receptor.model <- function(t, y, parms){
  with(as.list(c(parms, y)),{
    delta.R.m0 <- -k.d_Rm * R.m0
    delta.R <- k.s_R * R.m0 + R.f * k.re * DRN - k.on * D * R.0 - k.d_R * R.0
    delta.DR <- k.on * D * R.0 - k.T * DR
    delta.DRN <- k.T * DR - k.re * DRN
    return( list( c(delta.R.m0, delta.R, delta.DR, delta.DRN ) ) )
  })
}

changing.plots <- function(plot.value, ref.data, changed.data){
  # Create colours for the different lines (except for reference)
  colours <- hue_pal()(length(changed.data))
  data.names <- names(changed.data)
  # y.val inserts the plot.value for the corresponding row of data.values
  y.val <- data.values[plot.value,]
  # The plot
  plt <- ggplot(data = ref.data, mapping = aes(x = time, y = !!sym(y.val$name) ) ) +
           # Lines (reference data stays black)
           geom_line(aes(color = "Ref")) +
           unlist( mapply(function(single.data, data.name)
                      geom_line(data = single.data, linetype = "dashed",
                                aes(color = data.name ) ),
                    changed.data, data.names ) ) +
           # Labels
           labs(x = "Time", y = paste(y.val$ylabel, y.val$unit), title = y.val$title) +
           theme(legend.position = "bottom") +
           # Legend data
           scale_colour_manual(values = c("black", colours),
                               limits = c("Ref", names(changed.data) ) ) +
           # Legend correction
           guides(color = guide_legend(
                   title = "",
                   override.aes = list(linetype = c(1, rep(2, length(changed.data)))))
           )
  return(plt)
}

## The model with a dose of 20
params$D <- calc.D(20)
model_20 <- ode(times = times, y = state,
                 parms = params, func = grd_model, method = "euler")
model_20 <- as.data.frame(model_20)

## The model without drug influence
model_noDrugs <- as.data.frame(ode(times = times, y = state,
                 parms = params, func = receptor.model, method = "euler"))

data.list <- list("No drugs" = model_noDrugs)

plots <- lapply(c("R.m0", "R.0"), changing.plots, model_20, data.list)
arrange.plots(plots, "No Glucocorticoid receptor Regulation")
```
When the DR(N) has no influence on the mRNA~R~ synthesis, a big part of the first equation disappears: $k_{s_Rm}\cdot(1-\frac{DR(N)}{IC_{50_Rm}+DR(N)})$
The figures show, when DR(N) has no influence on the mRNA~R~ synthesis, the mRNA~R~ concentration drops to zero fairly quickly and stays zero due to no mRNA~R~ synthesis.
The blue line details the normal scenario in which the drug is working correctly with a dose of 20.

\newpage

## Stopping drug treatment at Steady State

```{r steady-model}
## Define new model
steady_model <- function(t, y, parms){
  with(as.list(c(parms, y)),{
    delta.R.m0 <- k.s_Rm * (1 - ( DRN / (IC.50_Rm + DRN) ) ) - k.d_Rm * R.m0
    delta.R.0 <- k.s_R * R.m0 + R.f * k.re * DRN - k.on * D * R.0 - k.d_R * R.0
    delta.DR <- k.on * D * R.0 - k.T * DR
    delta.DRN <- k.T * DR - k.re * DRN
    delta.D <- 0
    return( list( c(delta.R.m0, delta.R.0, delta.DR, delta.DRN, delta.D) ) )
  })
}

params_steady <- c(k.s_Rm = 2.90, IC.50_Rm = 26.2, k.on = 0.00329, k.T = 0.63,
                   k.re = 0.57, R.f = 0.49, k.d_R = 0.0572, k.d_Rm = 0.612, k.s_R = 3.22)

state_steady <- c(R.m0 = 4.74, R.0 = 267, DR = 0, DRN = 0, D = calc.D(20))

## Create the event for the model
trigger <- function(t, y, params){
  y["D"] <- 0
  return(y)
}

## Check when the steady state is found
root <- function (t, y, params){
  x <- unlist(grd_model(t, y, params))
  numb1 <- sum(abs(x)) - 1e-4
  numb2 <- numb1 + y["D"]

  return(c(numb1, numb2))
}

## Run the model with the turnover
steadys <- ode( times = times, y = state_steady, parms = params_steady,
                func = steady_model, rootfun = root,
                events = list(func = trigger, root = TRUE, terminalroot = 2) )

last.root <- attributes(steadys)$troot
steadys <- as.data.frame(steadys)

data.list <- list("Turnover" = steadys)
plots <- lapply(c("R.m0", "R.0", "DR", "DRN"), changing.plots, model_20,data.list)
arrange.plots(plots, expression("Turnover at steady state"))
```

\newpage

## Changes in k~on~ and k~re~

```{r kondiff}
times <- seq(0, 168, by = 0.1)

state <- c(R.m0 = 4.74, R.0 = 267, DR = 0, DRN = 0)

params <- c(k.s_Rm = 2.90, IC.50_Rm = 26.2, k.on = 0.00329, k.T = 0.63, k.re = 0.57,
            R.f = 0.49, k.d_R = 0.0572, k.d_Rm = 0.612, k.s_R = 3.22, D = calc.D(20))

k.on_models <- list("k/5" = 0.0039/5,
                    "k/2" = 0.0039/2,
                    "k*2" = 0.0039*2,
                    "k*5" = 0.0039*5)

for (i in seq_along(k.on_models)){
  params$k.on <- k.on_models[[i]]
  k.on_models[[i]] <- as.data.frame(ode(times = times, y = state, parms = params,
                                        func = grd_model, method = "euler"))
}

plots <- lapply(c("R.m0", "R.0", "DR", "DRN"), changing.plots, model_20, k.on_models)
arrange.plots(plots, expression("Different k"["on"]*" values"))
```

\newpage

```{r krediff}
k.re_models <- list("k/5" = 0.57/5,
                    "k/2" = 0.57/2,
                    "k*2" = 0.57*2,
                    "k*5" = 0.57*5)

for (i in seq_along(k.re_models)){
  params$k.re <- k.re_models[[i]]
  k.re_models[[i]] <- as.data.frame(ode(times = times, y = state, parms = params,
                                        func = grd_model, method = "euler"))
}

plots <- lapply(c("R.m0", "R.0", "DR", "DRN"), changing.plots, model_20, k.re_models)
arrange.plots(plots, expression("Different k"["re"]*" values"))
```

\newpage

## Block Synthesis of Receptor
To simulate this situation, k~s_R~ needs to be put to zero and R~f~ and k~re~ should output no influence on \Delta R. This changes the model a bit:

```{r receptor_block}
block_model <- function(t, y, parms){
  with(as.list(c(parms, y)),{
    delta.R.m0 <- k.s_Rm * (1 - ( DRN / (IC.50_Rm + DRN) ) ) - k.d_Rm * R.m0
    delta.R.0 <- k.s_R * R.m0 - k.on * D * R.0 - k.d_R * R.0
    delta.DR <- k.on * D * R.0 - k.T * DR
    delta.DRN <- k.T * DR - k.re * DRN
    return( list( c(delta.R.m0, delta.R.0, delta.DR, delta.DRN ) ) )
  })
}

params <- c(k.s_Rm = 2.90, IC.50_Rm = 26.2, k.on = 0.00329, k.T = 0.63, k.re = 0.57,
            R.f = 0.49, k.d_R = 0.0572, k.d_Rm = 0.612, k.s_R = 0, D = calc.D(20))
state <- c(R.m0 = 4.74, R.0 = 267, DR = 0, DRN = 0)
times <- seq(0, 168, by = 1)

no.Receptor <- as.data.frame(ode(times = times, y = state, parms = params,
                                 func = block_model, method = "euler"))

data.list <- list("No receptor" = no.Receptor)
plots <- lapply(c("R.m0", "R.0", "DR", "DRN"), changing.plots, model_20, data.list)
arrange.plots(plots, expression("No receptor synthesis"))
```

\newpage

## Increased and decreased baseline mRNA production
```{r baseline-change}
params <- c(k.s_Rm = 2.90, IC.50_Rm = 26.2, k.on = 0.00329, k.T = 0.63, k.re = 0.57,
            R.f = 0.49, k.d_R = 0.0572, k.d_Rm = 0.612, k.s_R = 0, D = calc.D(20))
state <- c(R.m0 = 4.74, R.0 = 267, DR = 0, DRN = 0)
times <- seq(0, 168, by = 0.1)

k.models <- list("k/5" = 2.9/5,
                 "k/2" = 2.9/2,
                 "k*2" = 2.9*2,
                 "k*5" = 2.9*5)

for (i in seq_along(k.models)){
  params$k.s_Rm <- k.models[[i]]
  params$k.d_Rm <- k.models[[i]]/4.74
  k.models[[i]] <- as.data.frame(ode(times = times, y = state, parms = params,
                                     func = grd_model, method = "euler"))
}

plots <- lapply(c("R.m0", "R.0", "DR", "DRN"), changing.plots, model_20, k.models)
arrange.plots(plots, expression("Different k"["s_Rm"]*" and k"["d_Rm"]*" values"))
```
